---
title: "Lab 1: Species Distribution Modeling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

#### Import Packages

```{r, warning = FALSE}
# load packages, installing if missing
if (!require(librarian)){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(
  dismo, dplyr, DT, ggplot2, here, htmltools, leaflet, mapview, purrr, raster, readr, rgbif, rgdal, rJava, sdmpredictors, sf, spocc, tidyr, GGally, caret, pdp, ranger, rpart, rpart.plot, rsample, skimr, vip, maptools, usdm)
select <- dplyr::select # overwrite raster::select
options(readr.show_col_types = FALSE,
        scipen = 999)

# set random seed for reproducibility
set.seed(42)

# directory to store data
dir_data      <- here("data/sdm")
env_stack_grd <- file.path(dir_data, "env_stack.grd")
mdl_maxv_rds  <- file.path(dir_data, "mdl_maxent_vif.rds")

dir.create(dir_data, showWarnings = F)

# read raster stack of environment
#env_stack <- raster::stack(env_stack_grd)

```

### Lab 1a: Exploratory Data Analysis

#### Get Species Observations

```{r, warning = FALSE}
obs_csv <- file.path(dir_data, "obs.csv")
obs_geo <- file.path(dir_data, "obs.geojson")
redo    <- FALSE

if (!file.exists(obs_geo) | redo){
  # get species occurrence data from GBIF with coordinates
  (res <- spocc::occ(
    query = 'Glaucomys sabrinus', 
    from = 'gbif', has_coords = T,
    limit = 10000))
  
  # extract data frame from result
  df <- res$gbif$data[[1]]
  readr::write_csv(df, obs_csv)
  
  # convert to points of observation from lon/lat columns in data frame
  obs <- df %>% 
    filter(latitude > 26, longitude > -160) %>%
    sf::st_as_sf(
      coords = c("longitude", "latitude"),
      crs = st_crs(4326)) %>% 
    select(prov, key) # save space (joinable from obs_csv)
  sf::write_sf(obs, obs_geo, delete_dsn=T)
}
obs <- sf::read_sf(obs_geo)
nrow(obs) # number of rows
```

```{r}
# show points on map
mapview::mapview(obs, map.types = "OpenTopoMap")
```

**Question 1. How many observations total are in GBIF for your species? (Hint: ?occ)**

*As we can see in the `df` dataframe there are 3797 observations total in GBIF for my species.*

**Question 2. Do you see any odd observations, like marine species on land or vice versa? If so, please see the Data Cleaning and explain what you did to fix or remove these points.**

*Yes, there are some strange observations that are in South America and Africa, when this species should only be seen in North America. So I filtered the dataset by lon and lat so that only oberservations in North America will show up. This leaves us with 3792 observations.*

#### Get Environmetal Data

##### Presence Points

```{r}
dir_env <- file.path(dir_data, "env")

# set a default data directory
options(sdmpredictors_datadir = dir_env)

# choosing terrestrial
env_datasets <- sdmpredictors::list_datasets(terrestrial = TRUE, marine = FALSE)

# show table of datasets
env_datasets %>% 
  select(dataset_code, description, citation) %>% 
  DT::datatable()
```

```{r}
# choose datasets for a vector
env_datasets_vec <- c("WorldClim", "ENVIREM")

# get layers
env_layers <- sdmpredictors::list_layers(env_datasets_vec)
DT::datatable(env_layers)
```

```{r, warning = FALSE}
# choose layers after some inspection and perhaps consulting literature
env_layers_vec <- c("WC_alt", "WC_bio1", "WC_bio2", "ER_tri", "ER_topoWet")

# get layers
env_stack <- load_layers(env_layers_vec)

# interactive plot layers, hiding all but first (select others)
# mapview(env_stack, hide = T) # makes the html too big for Github
plot(env_stack, nc=2)
```

```{r}
obs_hull_geo  <- file.path(dir_data, "obs_hull.geojson")
env_stack_grd <- file.path(dir_data, "env_stack.grd")

if (!file.exists(obs_hull_geo) | redo){
  # make convex hull around points of observation
  obs_hull <- sf::st_convex_hull(st_union(obs))
  
  # save obs hull
  write_sf(obs_hull, obs_hull_geo)
}
obs_hull <- read_sf(obs_hull_geo)

# show points on map
mapview(
  list(obs, obs_hull))
```

```{r, warning = FALSE}
if (!file.exists(env_stack_grd) | redo){
  obs_hull_sp <- sf::as_Spatial(obs_hull)
  env_stack <- raster::mask(env_stack, obs_hull_sp) %>% 
    raster::crop(extent(obs_hull_sp))
  writeRaster(env_stack, env_stack_grd, overwrite=T)  
}
env_stack <- stack(env_stack_grd)

# show map
# mapview(obs) + 
#   mapview(env_stack, hide = T) # makes html too big for Github
plot(env_stack, nc=2)
```
###### Psuedo-Absense Points

```{r, warning = FALSE}
absence_geo <- file.path(dir_data, "absence.geojson")
pts_geo     <- file.path(dir_data, "pts.geojson")
pts_env_csv <- file.path(dir_data, "pts_env.csv")


if (!file.exists(absence_geo) | redo){
  # get raster count of observations
  r_obs <- rasterize(
    sf::as_Spatial(obs), env_stack[[1]], field=1, fun='count')
  
  # show map
  # mapview(obs) + 
  #   mapview(r_obs)
  
  # create mask for 
  r_mask <- mask(env_stack[[1]] > -Inf, r_obs, inverse=T)
  
  # generate random points inside mask
  absence <- dismo::randomPoints(r_mask, nrow(obs)) %>% 
    as_tibble() %>% 
    st_as_sf(coords = c("x", "y"), crs = 4326)
  
  write_sf(absence, absence_geo, delete_dsn=T)
}
absence <- read_sf(absence_geo)

# show map of presence, ie obs, and absence
mapview(obs, col.regions = "green") + 
  mapview(absence, col.regions = "gray")
```

```{r, warning = FALSE}
if (!file.exists(pts_env_csv) | redo){

  # combine presence and absence into single set of labeled points 
  pts <- rbind(
    obs %>% 
      mutate(
        present = 1) %>% 
      select(present, key),
    absence %>% 
      mutate(
        present = 0,
        key     = NA)) %>% 
    mutate(
      ID = 1:n()) %>% 
    relocate(ID)
  write_sf(pts, pts_geo, delete_dsn=T)

  # extract raster values for points
  pts_env <- raster::extract(env_stack, as_Spatial(pts), df=TRUE) %>% 
    tibble() %>% 
    # join present and geometry columns to raster value results for points
    left_join(
      pts %>% 
        select(ID, present),
      by = "ID") %>% 
    relocate(present, .after = ID) %>% 
    # extract lon, lat as single columns
    mutate(
      #present = factor(present),
      lon = st_coordinates(geometry)[,1],
      lat = st_coordinates(geometry)[,2]) %>% 
    select(-geometry)
  write_csv(pts_env, pts_env_csv)
}
pts_env <- read_csv(pts_env_csv)

pts_env %>% 
  # show first 10 presence, last 10 absence
  slice(c(1:10, (nrow(pts_env)-9):nrow(pts_env))) %>% 
  DT::datatable(
    rownames = F,
    options = list(
      dom = "t",
      pageLength = 20))
```

#### Term Plots

```{r, warning = FALSE}
pts_env %>% 
  select(-ID) %>% 
  mutate(
    present = factor(present)) %>% 
  pivot_longer(-present) %>% 
  ggplot() +
  geom_density(aes(x = value, fill = present)) + 
  scale_fill_manual(values = alpha(c("gray", "green"), 0.5)) +
  scale_x_continuous(expand=c(0,0)) +
  scale_y_continuous(expand=c(0,0)) +
  theme_bw() + 
  facet_wrap(~name, scales = "free") +
  theme(
    legend.position = c(1, 0),
    legend.justification = c(1, 0))
```

### Lab 1b: Logistic Regression

### Explore (cont'd)

```{r}
# Look at the data in a dataframe.

datatable(pts_env, rownames = F)
```


```{r, warning = FALSE}
# Now look at it in a pairs plot.

GGally::ggpairs(
  select(pts_env, -ID),
  aes(color = factor(present)))
```

### Logistic Regression

#### Setup Data

```{r}
# setup model data
d <- pts_env %>% 
  select(-ID) %>%  # remove terms we don't want to model
  tidyr::drop_na() # drop the rows with NA values
nrow(d)
```

#### Linear Model

```{r}
# fit a linear model
mdl <- lm(present ~ ., data = d)
summary(mdl)
```

```{r}
y_predict <- predict(mdl, d, type="response")
y_true    <- d$present

range(y_predict)
```

```{r}
range(y_true)
```

#### Generalized Linear Model

```{r}
# fit a generalized linear model with a binomial logit link function
mdl <- glm(present ~ ., family = binomial(link="logit"), data = d)
summary(mdl)
```

```{r}
y_predict <- predict(mdl, d, type="response")

range(y_predict)
```

```{r}
# show term plots
termplot(mdl, partial.resid = TRUE, se = TRUE, main = F, ylim="free")
```

#### Generalized Additive Model

```{r}
librarian::shelf(mgcv)

# fit a generalized additive model with smooth predictors
mdl <- mgcv::gam(
  formula = present ~ s(WC_alt) + s(WC_bio1) + 
    s(WC_bio2) + s(ER_tri) + s(ER_topoWet) + s(lon) + s(lat), 
  family = binomial, data = d)
summary(mdl)
```

```{r}
# show term plots
plot(mdl, scale=0)
```

#### Maxent (Maximum Entropy)

```{r}
# load extra packages
librarian::shelf(
  maptools, sf)

mdl_maxent_rds <- file.path(dir_data, "mdl_maxent.rds")

# show version of maxent
if (!interactive())
  maxent()
```

```{r}
# get environmental rasters
# NOTE: the first part of Lab 1. SDM - Explore got updated to write this clipped environmental raster stack
env_stack_grd <- file.path(dir_data, "env_stack.grd")
env_stack <- stack(env_stack_grd)
plot(env_stack, nc=2)
```

```{r}
# get presence-only observation points (maxent extracts raster values for you)
obs_geo <- file.path(dir_data, "obs.geojson")
obs_sp <- read_sf(obs_geo) %>% 
  sf::as_Spatial() # maxent prefers sp::SpatialPoints over newer sf::sf class

# fit a maximum entropy model
if (!file.exists(mdl_maxent_rds)){
  mdl <- maxent(env_stack, obs_sp)
  readr::write_rds(mdl, mdl_maxent_rds)
}
mdl <- read_rds(mdl_maxent_rds)

# plot variable contributions per predictor
plot(mdl)
```

```{r}
# plot term plots
response(mdl)
```

```{r}
# predict
y_predict <- predict(env_stack, mdl) #, ext=ext, progress='')

plot(y_predict, main='Maxent, raw prediction')
data(wrld_simpl, package="maptools")
plot(wrld_simpl, add=TRUE, border='dark grey')
```

### Lab 1c: Decision Trees

```{r}
# graphical theme
ggplot2::theme_set(ggplot2::theme_light())

# read data
pts_env <- read_csv(pts_env_csv)
d <- pts_env %>% 
  select(-ID) %>%                   # not used as a predictor x
  mutate(
    present = factor(present)) %>%  # categorical response
  na.omit()                         # drop rows with NA
skim(d)
```

#### Split data into Training and Testing

```{r}
# create training set with 80% of full data
d_split  <- rsample::initial_split(d, prop = 0.8, strata = "present")
d_train  <- rsample::training(d_split)

# show number of rows present is 0 vs 1
table(d$present)
```

```{r}
table(d_train$present)
```

#### Decision Trees

##### Partition, depth = 1

```{r}
# run decision stump model
mdl <- rpart(
  present ~ ., data = d_train, 
  control = list(
    cp = 0, minbucket = 5, maxdepth = 1))
mdl
```

```{r}
# plot tree 
par(mar = c(1, 1, 1, 1))
rpart.plot(mdl)
```

##### Partition, depth = default

```{r}
# decision tree with defaults
mdl <- rpart(present ~ ., data = d_train)
mdl
```

```{r}
rpart.plot(mdl)

# plot complexity parameter
plotcp(mdl)

# rpart cross validation results
mdl$cptable
```

##### Feature Interpretation

```{r}
# caret cross validation results
mdl_caret <- train(
  present ~ .,
  data       = d_train,
  method     = "rpart",
  trControl  = trainControl(method = "cv", number = 10),
  tuneLength = 20)

ggplot(mdl_caret)
```

```{r}
vip(mdl_caret, num_features = 40, bar = FALSE)
```

```{r}
# Construct partial dependence plots
p1 <- partial(mdl_caret, pred.var = "lat") %>% autoplot()
p2 <- partial(mdl_caret, pred.var = "WC_bio2") %>% autoplot()
p3 <- partial(mdl_caret, pred.var = c("lat", "WC_bio2")) %>% 
  plotPartial(levelplot = FALSE, zlab = "yhat", drape = TRUE, 
              colorkey = TRUE, screen = list(z = -20, x = -60))

# Display plots side by side
gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
```

#### Random Forest

##### Fit

```{r}
# number of features
n_features <- length(setdiff(names(d_train), "present"))

# fit a default random forest model
mdl_rf <- ranger(present ~ ., data = d_train)

# get out of the box RMSE
(default_rmse <- sqrt(mdl_rf$prediction.error))
```

##### Feature Interpretation

```{r}
# re-run model with impurity-based variable importance
mdl_impurity <- ranger(
  present ~ ., data = d_train,
  importance = "impurity")

# re-run model with permutation-based variable importance
mdl_permutation <- ranger(
  present ~ ., data = d_train,
  importance = "permutation")
```

```{r}
p1 <- vip::vip(mdl_impurity, bar = FALSE)
p2 <- vip::vip(mdl_permutation, bar = FALSE)

gridExtra::grid.arrange(p1, p2, nrow = 1)
```

### Lab 1d: Evaluate Models

```{r}
# read points of observation: presence (1) and absence (0)
pts <- read_sf(pts_geo)
```

#### Split observations into Training and Testing

```{r}
# create training set with 80% of full data
pts_split  <- rsample::initial_split(
  pts, prop = 0.8, strata = "present")
pts_train  <- rsample::training(pts_split)
pts_test   <- rsample::testing(pts_split)

pts_train_p <- pts_train %>% 
  filter(present == 1) %>% 
  as_Spatial()
pts_train_a <- pts_train %>% 
  filter(present == 0) %>% 
  as_Spatial()
```

#### Calibrate: Model Selection

```{r}
# show pairs plot before multicollinearity reduction with vifcor()
pairs(env_stack)
```

```{r}
# calculate variance inflation factor per predictor, a metric of multicollinearity between variables
vif(env_stack)
```

```{r}
# stepwise reduce predictors, based on a max correlation of 0.7 (max 1)
v <- vifcor(env_stack, th=0.7) 
v
```

```{r}
# reduce enviromental raster stack by 
env_stack_v <- usdm::exclude(env_stack, v)

# show pairs plot after multicollinearity reduction with vifcor()
pairs(env_stack_v)
```

```{r}
# fit a maximum entropy model
if (!file.exists(mdl_maxv_rds)){
  mdl_maxv <- maxent(env_stack_v, sf::as_Spatial(pts_train))
  readr::write_rds(mdl_maxv, mdl_maxv_rds)
}
mdl_maxv <- read_rds(mdl_maxv_rds)

# plot variable contributions per predictor
plot(mdl_maxv)
```

```{r}
# plot term plots
response(mdl_maxv)
```

```{r}
# predict
y_maxv <- predict(env_stack, mdl_maxv) #, ext=ext, progress='')

plot(y_maxv, main='Maxent, raw prediction')
data(wrld_simpl, package="maptools")
plot(wrld_simpl, add=TRUE, border='dark grey')
```

#### Evaluate: Model Performance

##### Area Under the Curve (AUC), Reciever Operater Characteristic (ROC) Curve and Confusion Matrix  

```{r}
pts_test_p <- pts_test %>% 
  filter(present == 1) %>% 
  as_Spatial()
pts_test_a <- pts_test %>% 
  filter(present == 0) %>% 
  as_Spatial()

y_maxv <- predict(mdl_maxv, env_stack)
#plot(y_maxv)

e <- dismo::evaluate(
  p     = pts_test_p,
  a     = pts_test_a, 
  model = mdl_maxv,
  x     = env_stack)
e
```

```{r}
plot(e, 'ROC')

thr <- threshold(e)[['spec_sens']]
thr
```

```{r}
p_true <- na.omit(raster::extract(y_maxv, pts_test_p) >= thr)
a_true <- na.omit(raster::extract(y_maxv, pts_test_a) < thr)

# (t)rue/(f)alse (p)ositive/(n)egative rates
tpr <- sum(p_true)/length(p_true)
fnr <- sum(!p_true)/length(p_true)
fpr <- sum(!a_true)/length(a_true)
tnr <- sum(a_true)/length(a_true)

matrix(
  c(tpr, fnr,
    fpr, tnr), 
  nrow=2, dimnames = list(
    c("present_obs", "absent_obs"),
    c("present_pred", "absent_pred")))
```

```{r}
# add point to ROC plot
plot(e, 'ROC')
points(fpr, tpr, pch=23, bg="blue")
```

```{r}
plot(y_maxv > thr)
```






